---
title: "Class 13: RNAseq mini project"
author: "Kelsey Fierro"
format: gfm
---

## Background

Today we will run through a complete RNASeq analysis 

The data for for hands-on session comes from GEO entry: GSE37704, which is associated with the following publication:

Trapnell C, Hendrickson DG, Sauvageau M, Goff L et al. "Differential analysis of gene regulation at transcript resolution with RNA-seq". Nat Biotechnol 2013 Jan;31(1):46-53. PMID: 23222703

## Data Import
```{r}
counts <- read.csv("GSE37704_featurecounts.csv",row.names = 1)
metadata <- read.csv("GSE37704_metadata.csv")
```

check correspondance of 'metadata' and 'counts' (ie that the columns in )
```{r}
head(counts)
```

```{r}
head(metadata)
```
```{r}
metadata$id
```

Fix to remove that first "length" column of 'counts', if this first column is not out then there will be issues later!!
```{r}
counts <- as.matrix(counts[,-1])
```

remove the low counts

```{r}
tot.counts <- rowSums(counts)
head(tot.counts)
```
```{r}
zero.inds <- tot.counts == 0
head(zero.inds)

```
```{r}
counts <- counts[!zero.inds,]
#counts <- as.matrix(counts[,-1])
```


```{r}
test_cols <- all(colnames(counts)[-1] == metadata$id)
```



```{r}
if( test_cols ) {
  message("Wow... there is a problem with the metadata counts setup!")
}
```
```{r}
head(counts)
```


## Setup for DESeq

```{r}
library(DESeq2)
```

```{r}
dds <- DESeqDataSetFromMatrix(countData = counts,
                              colData = metadata,
                              design = ~condition)
```

## Run DESeq

```{r}
dds <- DESeq(dds)
```


## Get Results

```{r}
res <- results(dds)
```

## Add Annotation

```{r}
head(res)
```

## Visualize results

```{r}
plot( res$log2FoldChange, -log(res$padj))
```
```{r}
library(ggplot2)

ggplot(res) +
  aes(log2FoldChange, -log(padj)) +
  geom_point() +
  geom_vline(xintercept = c(-2,2), col="red") +
  geom_hline(yintercept = -log(0.05), col="red")
```

```{r}
my_cols <- rep("gray", nrow(res))
ggplot(res) +
  aes(log2FoldChange, -log(padj)) +
  geom_point(col=my_cols) +
  geom_vline(xintercept = c(-2,2), col="red") +
  geom_hline(yintercept = -log(0.05), col="red")
```
```{r}
my_cols[ abs(res$log2FoldChange) > 2 ] <-  "blue"
ggplot(res) +
  aes(log2FoldChange, -log(padj)) +
  geom_point(col=my_cols) +
  geom_vline(xintercept = c(-2,2), col="red") +
  geom_hline(yintercept = -log(0.05), col="red")
```
```{r}
my_cols[abs(res$log2FoldChange) >= 2] <- "blue"
res$padj >= 0.05
ggplot(res) +
  aes(log2FoldChange, -log(padj)) +
  geom_point(col=my_cols) +
  geom_vline(xintercept = c(-2,2), col="red") +
  geom_hline(yintercept = -log(0.05), col="red")
```


## Pathway analysis

```{r}
library(gage)
library(gageData)

data(kegg.sets.hs)
data(sigmet.idx.hs)
```
```{r}
#kegs.sets.hs = kegg.sets.hs[sigment.idx.hs]
head(kegg.sets.hs, 3)
```
```{r}
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```
```{r}
keggres = gage(foldchanges, gsets=kegg.sets.hs)
attributes(keggres)
```
```{r}
head(keggres$less)
```
```{r}
library(pathview)
pathview(gene.data=foldchanges, pathway.id="hsa04110")
```
# Lets try GO analysis

```{r}
data(go.sets.hs)
data(go.subs.hs)

# Focus on Biological Process subset of GO
gobpsets <-  go.sets.hs[go.subs.hs$BP]

gobpres <-  gage(foldchanges, gsets=gobpsets, same.dir=TRUE)

lapply(gobpres, head)
```
```{r}
head( gobpres$less)
```
# Reactome

some folks really like ractome online (ie their webpage viewer < (https://reactome.org/) ) rather than the R package of the same name (available from bioconductor)

To use the website viewer we want to upload our set of gene symbols for the genes we want to focus on (here those with a P-value below 0.05)

```{r}
#sig_genes <- res[res$padj <- 0.05 & !is.na(res$padj), "symbol"]
#write.table(sig_genes, file="significant_genes.txt",
            #row.names=FALSE, col.names=FALSE, quote=FALSE)
```


## Save results

```{r}
write.csv(res, file="myresults.csv")
```

```{r}
save(res, counts, metadata, file="my_results.RData")
```

