---
title: "class10"
author: "Kelsey Fierro"
format: pdf
---

```{r}
candy <- read.csv("candy-data.csv", row.names = 1)
```
```{r,, message=FALSE}
library(dplyr)
```

> Q1. How many different candy types are in this dataset?

```{r}
dim(candy)
```
13 types of candy!

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

For M&M's:
```{r}
candy$winpercent[34]
```


> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy$winpercent[29]
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy$winpercent[78]
```


> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

winpercent seems to be the odd column out in terms of scale and measurement.

> Q7. What do you think a zero and one represent for the candy$chocolate column?

It represents whether or not the candy has chocolate in it or not

> Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)
ggplot(candy) +
  aes(winpercent) +
  geom_histogram(bins=20)
```


> Q9. Is the distribution of winpercent values symmetrical?

```{r}
ggplot(candy) +
  aes(winpercent) +
  geom_density()
```
No the distribution is not symmetrical

> Q10. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
```
```{r}
summary(candy$winpercent)
```

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
#find all chocolate candy in dataset, extract their winpercent values, find mean of values, then do same for fruity candy and find which ones higher

choc.inds <- as.logical(candy$chocolate)
choc.candy <- candy[choc.inds,]
choc.win <- choc.candy$winpercent
choc.mean <- mean(choc.win)
choc.mean
```
```{r}
fruit.inds <- as.logical(candy$fruity)
fruit.candy <- candy[fruit.inds,]
fruit.win <- fruit.candy$winpercent
fruit.mean <- mean(fruit.win)
fruit.mean
```
```{r}
choc.mean>=fruit.mean
```
Chocolate is ranked higher than fruit

> Q12. Is this difference statistically significant?

```{r}
t.test(choc.win, fruit.win)
```

>Q13. What are the five least liked candy types in this set?

```{r}
ord.ind <- order(candy$winpercent)
head(candy[ord.ind, ], 5)
```

>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
tail(candy[ord.ind, ], 5)
```

>Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()
```
> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col()
```
```{r}
mycols <- rep("black", nrow(candy))
mycols[candy$chocolate==1] <- "chocolate"
mycols[candy$bar==1] <- "brown"
mycols[candy$fruity==1] <- "pink"

ggplot(candy) +
  aes(x=winpercent, 
      y=reorder( rownames(candy), winpercent)) +
  geom_col(fill=mycols)
```
> Q17. What is the worst ranked chocolate candy?

The worst ranked candy is Sixlets.

> Q18. What is the best ranked fruity candy?

The best ranked fruity candy is Starburst! 


# Winpercent vs pricepoint

```{r}
library(ggrepel)
library(ggplot2)
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=mycols) +
  geom_text(col=mycols) +
  geom_text_repel(col=mycols, size=3.3, max.overlaps=50)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's miniatures offer the best bang for your buck!

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord1 <- order(candy$pricepercent, decreasing = TRUE)
head(candy[ord1,c(11,12)], n=5)
```
Nik L Nip is tied for most expensive but is somehow one of the least popular candies overall.

> Q21. Optional

```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_col()
```
```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent),
                   xend = 0), col="gray40") +
  geom_point()
```


# correlation

```{r}
library(corrplot)
cor(candy)
cij <- cor(candy)
corrplot(cij)
```

>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and fruity are anti-correlated, as well as pluribus and bar.

>Q23. Similarly, what two variables are most positively correlated?

Winpercent and chocolate, and bar and chocolate are all positively correlated.


# principal component analysis
the main function in base R for this 'prcomp()' and we want to set 'scale=TRUE'

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```
```{r}
#pca$x
library(ggrepel)
ggplot(pca$x) +
  aes(PC1, PC2, label=rownames(pca$x)) +
  geom_point(col=mycols) +
  geom_text_repel(col=mycols)
```

dont forget about your variable "loadings" - how the original variables contribute to your new PCs ...

```{r}
ggplot(pca$rotation) +
  aes(PC1, rownames(pca$rotation)) +
  geom_col()
```
> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

PC1 has the fruity and pluribus variables in a strong positive direction.  This makes sense as many fruity candies are small pieces of candy that are packaged together in a bag.
